Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder

### 2024-04-30 5_artigos/statistical-modelling/smj/paper-lineu.tex
inclui artigo smj

\documentclass[a4paper]{article}

\usepackage[pages=all, color=black, position={current page.south}, placement=bottom, scale=1, opacity=1, vshift=5mm]{background}
%\SetBgContents{
%	\tt This work is shared under a \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC BY-SA 4.0 license} unless otherwise noted
%}      % copyright

\usepackage[margin=1in]{geometry} % full-width

% AMS Packages
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

% Unicode
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\hypersetup{
	unicode,
%	colorlinks,
%	breaklinks,
%	urlcolor=cyan, 
%	linkcolor=blue, 
	%pdfauthor={Author One, Author Two, Author Three},
	%pdftitle={A simple article template},
	%pdfsubject={A simple article template},
	%pdfkeywords={article, template, simple},
	%pdfproducer={LaTeX},
	%pdfcreator={pdflatex}
}

% Vietnamese
%\usepackage{vntex}

% Natbib
\usepackage[sort&compress,numbers,square]{natbib}
\bibliographystyle{mplainnat}

% Theorem, Lemma, etc
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}{Claim}[theorem]
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{hypothesis}[theorem]{Hypothesis}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{criterion}[theorem]{Criterion}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{principle}[theorem]{Principle}

\usepackage{graphicx, color}
\graphicspath{{fig/}}

%\usepackage[linesnumbered,ruled,vlined,commentsnumbered]{algorithm2e} % use algorithm2e for typesetting algorithms
\usepackage{algorithm, algpseudocode} % use algorithm and algorithmicx for typesetting algorithms
\usepackage{mathrsfs} % for \mathscr command

\usepackage{multirow}
\usepackage{float}

% Author info
\title{Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.}
\author{Lineu Alberto Cavazani de Freitas$^1$\thanks{Author One was partially supported by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001.} \and Ligia de Oliveira Carlos$^2$ \and Antônio Carlos Ligocki Campos$^2$ \and Wagner Hugo Bonat$^3$}

\date{
	$^1$Department of Informatics, Paraná Federal University, Paraná, Brazil\\%
	$^2$Department of Surgical Clinic, Paraná Federal University, Paraná, Brazil\\%
        $^3$Department of Statistics, Paraná Federal University, Paraná, Brazil\\%
%	\today
}

\begin{document}
	\maketitle

	\begin{abstract}

\noindent Clinical trials are common in medical research where multiple non-Gaussian responses and time-dependent observations are frequent.
The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. 
We propose a general strategy based on the Wald statistics to perform hypothesis tests like ANOVAs, MANOVAs and multiple comparison tests on 
regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). 
McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. 
We design different simulation scenarios to verify the properties of the proposed tests. 
The results are promising showing that the proposed tests present the levels of confidence close to the specified one for all simulation study scenarios.
Complementary to the proposal, we developed implementations in the R language to carry out the tests presented, the codes are available in the supplementary material.
The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. 
The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. 
    The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.

\vspace{0.5cm}
		\noindent\textbf{Keywords:} McGLM, Hypothesis tests, Wald test, ANOVA, MANOVA, Multiple comparisons
	\end{abstract}

	%\tableofcontents

@@

According to Meinert and Tonascia \cite{meinert1986clinical} ``a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

In this type of study participants are randomly assigned to a treatment or control group (or to multiple treatment groups) \cite{hannan2008randomized} and the objective is to assess whether there is a significant difference between the groups.

@@

Often, in medical studies that use clinical trials, multiple outcomes are taken for a group of patients, for example: Kangovi et al \cite{kangovi2018effect} evaluated the effect of community health worker support on clinical outcomes of low-income patients; Song and Baicker \cite{song2019effect} studied the effect of a workplace wellness program on employee health and economic outcomes; Thyregod et al \cite{thyregod2019five} presents the results of a five-year clinical and echocardiographic outcomes from the nordic aortic valve intervention (NOTION); Schmitz et al \cite{schmitz2019effect} evaluated the effect of Home-Based Exercise and Weight Loss Programs on Breast Cancer–Related Lymphedema Outcomes; Bonat et al \cite{plastica} proposed a model to deal with multiple outcomes in repeated measures studies motivated by a problem that aimed to compare different aesthetic eyelid surgery techniques; Petterle et al \cite{petterle2021multivariate} proposed a multivariate regression model to handle multiple continuous bounded outcomes motivated by a body fat percentage data. Consequently, the joint analysis of multiple outcomes studies has been of increased interest in the medical and statistical literature.

@@

In this article, we adopt the multivariate covariance generalized linear models (McGLM) framework \cite{Bonat16}, which provides an environment for modeling multiple non-Gaussian responses simultaneously with flexible and interpretable modeling of the covariance structure. The within outcomes covariance matrix is specified for each marginal outcome using a linear combination of known matrices, while the joint covariance matrix is specified using the generalized Kronecker product \cite{martinez13,Bonat16}.

@@

When working with regression models, a common interest is to verify if the absence of a certain explanatory variable of the model would generate a loss in the adjustment. Thus, a conjecture of interest is to assess whether there is sufficient evidence in the data to state that a given explanatory variable has no effect on the response. This is done through hypothesis testing. In this context, three hypothesis tests are common: the likelihood ratio test, the Wald test, and the Lagrange multiplier test, also known as the score test. Engle \cite{engle} describes the general formulation of the three tests. All of them are based on the likelihood function of the models.

@@

In the case of standard linear models, there are techniques such as analysis of variance (ANOVA), initially proposed by Fisher and Mackenzie \cite{anova_fisher}. According to St, Wold et al \cite{anova1}, ANOVA is one of the most widely used statistical methods to test hypotheses and it is present in virtually all introductory statistics materials. The objective of the technique is to evaluate the effect of each explanatory variables on the response. This is done by comparing models with and without each of the explanatory variables. Therefore, this procedure makes it possible to assess whether the removal of each of the variables generates a significantly worse model when compared to the model with the variable. For the multivariate case, the technique of analysis of variance (ANOVA) is extended to the multivariate analysis of variance \citep{manova}, MANOVA. Among the multivariate hypothesis tests already discussed in the literature for MANOVAs, we have the Wilk's \cite{wilks} $\lambda$, Hotelling-Lawley trace \cite{lawley,hotelling}, Pillai trace \cite{pillai} and largest root of Roy \cite{roy}.

@@

Complementary to ANOVAs and MANOVAs are multiple comparison tests. Such procedures are used when the analysis of variance points to the existence of a significant effect of the parameters associated with a categorical variable, that is, there is at least one significant difference between the levels of a factor. Thus, the multiple comparison test is another procedure based on hypothesis testing, used to determine where these differences appear. For example, suppose there is a three-level categorical variable $X$ in the model that assumes the levels: A, B, or C. The analysis of variance will show if there is an effect of the variable $X$ in the model, that is, if the response values are associated with the levels of $X$, however this result will not tell us whether the response values differ from A to B, or from A to C, or whether B differs from C. To detect such differences, multiple comparison tests are used, some of the tests discussed in the literature are: the Dunnett, Tukey, student's t test (LSD), Scott-Knott test, among others. Hsu \cite{hsu1996multiple} discusses various procedures for the purpose of multiple comparisons. Bretz, Hothorn, and Westfall \cite{bretz2008multiple} presents procedures for multiple comparisons in linear models.

@@

It is important to note that the asymptotic normality of estimators of McGLMs was established in Bonat and Jørgensen \cite{Bonat16}, and Wald testing for estimators which are asymptotically normal is a standard approach in asymptotic statistics due to the central limit theorem. So, in this work, our objective is to show how to proceed to perform Wald tests for McGLMs and verify if the expected properties are met.

@@

We present the dataset in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of Bonat and Jørgensen \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model and present the main results of the analysis of the data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations. The data sets and R codes are available online and can be assessed in \href{https://github.com/lineu96/paper-htmcglm}{https://github.com/lineu96/paper-htmcglm}.

@@

The evaluation of binge eating was based on the binge eating scale (BES), one of the most used tools to measure binge eating in research, with numerous studies proving its effectiveness, translated into Portuguese and validated for individuals with obesity and submitted to bariatric surgery. This is a 16-item Likert scale questionnaire, prepared according to the Diagnostic and Statistical Manual of Mental Disorders (3rd edition) \cite{spitzer1980diagnostic} by Gormally et al \cite{gormally1982assessment}. The individuals were instructed to select the option that best represented their answer and the final score was the sum of the points for each item, this score ranging from 0 to 46.

To assess food addiction, the Food Addiction Scale (YFAS) was used, a questionnaire that seeks to detect symptoms of addictive eating behaviors. YFAS was based on substance dependence criteria from Diagnostic and Statistical Manual IV – Proofreading (DSM-IV-TR) \cite{segal2010diagnostic} and endorsed for highly processed foods. This questionnaire was developed by Gearhardt et al \cite{gearhardt2009preliminary} and is a combination of 25 Likert scale options and the evaluation option used was the number of addiction symptoms.

@@

\noindent where the $g_r()$ are standard link functions that relates the response to the predictor; $\boldsymbol{X}_r$ denotes a $N \times k_r$ design matrix; $\boldsymbol{\beta}_r$ denotes a  $k_r \times 1$ vector of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ denotes the generalized Kronecker product \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

The power variance function characterizes the Tweedie family of distributions, it is given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_r}_r$, in which some distributions stand out: Normal ($p $ = 0), Poisson ($p$ = 1), gamma ($p$ = 2) and Inverse Gaussian ($p$ = 3) \cite{Jorgensen87, Jorgensen97}.

The Poisson–Tweedie dispersion function \cite{Jorgensen15} is an option for responses that characterize counts. The dispersion function given by $\vartheta\left(\cdot; p\right) = \mu + \tau\mu^p$ where $\tau$ is the dispersion parameter. Thus, we have a rich class of models to deal with responses that characterize counts, since many important distributions appear as special cases, such as: Hermite ($p$ = 0), Neyman type A ($p$ = 1), negative binomial ($p$ = 2) and Poisson–inverse Gaussian (p = $3$).

@@

The dispersion matrix $\boldsymbol{\Omega({\tau})}$ describes the part of the covariance within each response variable that does not depend on the mean structure, ie, the correlation structure between the sample observations. Based on the ideas of Anderson et al \cite{Anderson73} and Pourahmadi \cite{Pourahmadi00}, Bonat and Jørgensen \cite{Bonat16} proposed to model the dispersion matrix through a matrix linear predictor combined with a covariance link function given by:

@@

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by Pinheiro and Bates \cite{Pinheiro96} and it is possible to select combinations of matrices to obtain the most well-known models in the literature for longitudinal data, time series, spatial and spatiotemporal data. Further details are discussed by Demidenko \cite{Demidenko13}.

@@

McGLMs are fitted based on the estimating function approach described in detail by Bonat and Jørgensen \cite{Bonat16} and Jørgensen and Knudsen \cite{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

@@

To estimate the regression parameters, the following quasi-score function \cite{Liang86} is used, 
$$
\begin{aligned}
  \psi_{\boldsymbol{\beta}}(\boldsymbol{\beta},
  \boldsymbol{\lambda}) = \boldsymbol{D}^\top
  \boldsymbol{C}^{-1}(\mathcal{Y} - \mathcal{M}),
\end{aligned}
$$

@@

To take into account the covariance between the vectors $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$, Bonat and Jørgensen \cite{Bonat16} obtained the sensitivity and cross-variability matrices, denoted by $S_{\boldsymbol{ \lambda \beta}}$, $S_{\boldsymbol{\beta \lambda}}$ and $V_{\boldsymbol{\lambda \beta}}$, more details in Bonat and Jørgensen \cite{Bonat16}. The joint sensitivity and variability matrices of $\psi_{\boldsymbol{\beta}}$ and $\psi_{\boldsymbol{\lambda}}$ are denoted by

@@

To solve the system of equations $\psi_{\boldsymbol{\beta}} = 0$ and $\psi_{\boldsymbol{\lambda}} = 0$, Jørgensen and Knudsen \cite{jorg04} proposed the modified Chaser algorithm, which is defined as

@@

We are interested in to assess whether there is sufficient evidence to state that there is an effect of the explanatory variable $X_1$ on both responses simultaneously. 
In this case, we have to test 2 parameters: $\beta_{11}$, which associates $X_1$ with the first response and $\beta_{21}$, which associates $X_1$ with the second response. 
We can write the hypothesis as follows:

@@

However, in cases where the responses have identical linear predictors and the assumptions about the parameters do not change from response to response, an alternative specification of the procedure is to use the Kronecker product as used in Bonat et al \cite{plastica}.

@@

Based on the proposal to use the Wald test for McGLMs, we aim to propose in this article three different procedures for generating ANOVA and MANOVA tables for regression parameters, following the nomenclature types I, II and III. In addition, we also aim to propose a procedure similar to ANOVA and MANOVA to evaluate the dispersion parameters of a given model. In the case of ANOVAs, a table is generated for each response variable. For MANOVAs only one tables is generated, therefore, in order to be able to perform MANOVAs, the model responses must be subject to the same linear predictor.

@@

When ANOVA shows a significant effect of a categorical variable, it is usually of interest to assess which of the levels differ from each other. For this, multiple comparison tests are used. In the literature there are several procedures to perform such tests, many of them described in Hsu \cite{hsu1996multiple}.

@@

To carry out a test of multiple comparisons, it is enough to select the desired contrasts in the lines of the matrix $\boldsymbol{K_1}$ and use these lines as a matrix for specifying the hypotheses of the Wald test. Finally, due to the amount of hypothesis testing in a multiple comparison procedure, correction of p-values by means of Bonferroni correction is recommended.

@@

We can opt for a type II multivariate analysis of variance to assess the importance of the variables in the problem. This result, presented in the table \autoref{tab4}, points to the significant existence of the moment effect and the absence of the group effect, indicating that for both responses the metrics change over time but without change between groups.

@@

\paragraph{Acknowledgements:} The authors thank the reviewers for their constructive and helpful comments, which greatly improved the article.

>>>>

%%%
%%%  LaTeX template for publications
%%%  to be submitted to Statistical Modelling
%%%
%%%  Prepared by Arnost Komarek
%%%  Version 0.2 (20140214)
%%%    0.2:  style of references slightly changed,
%%%          support for use with bibTeX added
%%%    0.3:  fax removed, orcid added, sections on supplementary material added as well as appendix
\documentclass[submit]{smj}

%%%%% PREAMBLE
%%%%% =============================================================================

%%% Place for putting personal \usepackage and \newcommand commands
%%% Note that some packages are loaded automatically
%%% with the smj class. 
%%% These include: graphicx, color, fancyvrb, amsmath, amssymb, calc, upquote (if available), natbib, url, hyperref.
%%%
%%% Please, specify all your personal definitions, newcommand etc. here
%%% and not inside the main body of the text.
%%% -------------------------------------------------------------------------------
\usepackage{multirow}
\usepackage{float}

%%% Identification of authors
%%% -------------------------------------------------------------------------------
%%% For each author, provide his/her first name, surname and possibly initials 
%%% of the middle names. 
%%%
%%% Use \Affil{NUMBER} following the author name for each unique affiliation,
%%% where NUMBER is integer starting from 1 to the number of affiliations needed
%%% in this paper. In case of multiple affiliations of one author, use
%%% \Affil{NUMBER1,}\Affil{NUMBER2,}\Affil{NUMBER3} following the author's name
%%% as it is done for Emmanuel Lesaffre below.

%%% The ORCID number does not have to be provided. If you want to provide it,
%%% put it in the \ORCID{} command while following (the last) \Affil{} command
%%% of given author as shown below.

  %%% For papers with 3 or more authors:
  %%%  in \Author{}, separate the authors with commas, the last author is separated by `\and' without a comma,
  %%%  in \AuthorRunning{}, use the full name of the first author followed by \textrm{et al.}.

\Author{Lineu Alberto Cavazani de Freitas\Affil{1}\ORCID{0000-0002-0076-6642}, 
        Ligia de Oliveira Carlos\Affil{2}\ORCID{0000-0002-2049-4592}, 
        Antônio Carlos Ligocki Campos\Affil{2}, 
        \and Wagner Hugo Bonat\Affil{3}\ORCID{0000-0002-0349-7054}
        %and Emmanuel Lesaffre\Affil{4,}\Affil{5}\ORCID{0000-0002-3747-6905}
}
\AuthorRunning{de Freitas \textrm{et al.}}

  %%% For papers with 2 authors:
  %%%  in both \Author{} and \AuthorRunning{},
  %%%  use the full names of both authors separated by '\and' without a comma.
%\Author{Arno\v{s}t Kom\'arek\Affil{1} \and Brian Marx\Affil{2}}
%\AuthorRunning{Arno\v{s}t Kom\'arek \and Brian Marx}

  %%% For papers with 1 author:
  %%%  in both \Author{} and \AuthorRunning{},
  %%%  use the full name the author.
%\Author{Arno\v{s}t Kom\'arek\Affil{1}}
%\AuthorRunning{Arno\v{s}t Kom\'arek}

%%% Affiliations as they should appear on the title page.
%%% -------------------------------------------------------------------------------
%%% Do not provide the full addresses here.
%%% The ordering inside \Affiliations{} should correspond to NUMBERs used 
%%% in \Affil{} commands in \Author{}
\Affiliations{

  %%% 1
\item Department of Informatics, 
      Paraná Federal University,
      Paraná,
      Brazil

  %%% 2
\item Department of Surgical Clinic, 
      Paraná Federal University,
      Paraná,
      Brazil

  %%% 3
\item Department of Statistics, 
      Paraná Federal University,
      Paraná,
      Brazil

  %%% 4
%\item Department of Biostatistics,
%      Erasmus University Rotterdam,
%      Rotterdam,
%      the Netherlands

  %%% 5
%\item Interuniversity Institute for Biostatistics and Statistical Bioinformatics,
%      KU Leuven and Universiteit Hasselt,
%      Leuven,
%      Belgium
}   %% end \Affiliations

%%% Postal, e-mail address, phone and fax of the corresponding author (not necessarily the first author).
%%% ------------------------------------------------------------------------------------------------------
%%% Use command \CorrAddress{} to provide a full postal address of the
%%% corresponding author in the form
%%% "Firstname Lastname, Department, University, Street 1, ZIP City, Country" 
%%% Use command \CorrEmail{} to provide an e-mail address of the corresponding author.
%%% Use command \CorrPhone{} to provide a phone number (including the country code!) of the corresponding author.
%%% Use command \CorrFax{} to provide a fax number (including the country code!) of the corresponding author.
\CorrAddress{Lineu Alberto Cavazani de Freitas, 
             Paraná Federal University, R. Evaristo F. Ferreira da Costa, s/n, Campus 3, R. Centro Politécnico - Jardim das Américas, Curitiba - PR
             }
\CorrEmail{lineuacf@gmail.com}
\CorrPhone{+55 41 9976-8582}

%%% Title and a short title (to be used as a running header) of the paper
%%% -------------------------------------------------------------------------------
\Title{Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder}
\TitleRunning{Template paper}

%%% Abstract
%%% -------------------------------------------------------------------------------
\Abstract{
Clinical trials are common in medical research where multiple non-Gaussian responses and time-dependent observations are frequent.
The analysis of data from these studies requires statistical modeling techniques that take these characteristics into account. 
We propose a general strategy based on the Wald statistics to perform hypothesis tests like ANOVAs, MANOVAs and multiple comparison tests on 
regression and dispersion parameters of multivariate covariance generalized linear models (McGLMs). 
McGLMs provide a general statistical modeling framework for normal and non-normal multivariate data analysis along with a wide range of correlation structures. 
We design different simulation scenarios to verify the properties of the proposed tests. 
The results are promising showing that the proposed tests present the levels of confidence close to the specified one for all simulation study scenarios.
Complementary to the proposal, we developed implementations in the R language to carry out the tests presented, the codes are available in the supplementary material.
The proposal is motivated by the analysis of a clinical trial that aims to evaluate the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. 
The subjects were separated into two groups (placebo and treatment) and evaluated at three different times. 
    The results indicate that addiction and binge eating disorder reduce over time, but there is no difference between groups at each time point.
}

%%% Key words
%%% -------------------------------------------------------------------------------
\Keywords{
McGLM; Hypothesis tests; Wald test; ANOVA; MANOVA; Multiple comparisons
}

%%%%% MAIN BODY 
%%%%% =============================================================================
\begin{document}

%%% Title page
%%% -------------------------------------------------------------------------------
%%% Use command\maketitle to produce the title page.
\maketitle

%%% Main text
%%% ------------------------------------------

@@

According to \cite{meinert1986clinical} ``a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

In this type of study participants are randomly assigned to a treatment or control group (or to multiple treatment groups) \citep{hannan2008randomized} and the objective is to assess whether there is a significant difference between the groups.

@@

Often, in medical studies that use clinical trials, multiple outcomes are taken for a group of patients, for example: \cite{kangovi2018effect} evaluated the effect of community health worker support on clinical outcomes of low-income patients; \cite{song2019effect} studied the effect of a workplace wellness program on employee health and economic outcomes;  \cite{thyregod2019five} presents the results of a five-year clinical and echocardiographic outcomes from the nordic aortic valve intervention (NOTION); \cite{schmitz2019effect} evaluated the effect of Home-Based Exercise and Weight Loss Programs on Breast Cancer–Related Lymphedema Outcomes; \cite{plastica} proposed a model to deal with multiple outcomes in repeated measures studies motivated by a problem that aimed to compare different aesthetic eyelid surgery techniques; \cite{petterle2021multivariate} proposed a multivariate regression model to handle multiple continuous bounded outcomes motivated by a body fat percentage data. Consequently, the joint analysis of multiple outcomes studies has been of increased interest in the medical and statistical literature.

@@

In this article, we adopt the multivariate covariance generalized linear models (McGLM) framework \citep{Bonat16}, which provides an environment for modeling multiple non-Gaussian responses simultaneously with flexible and interpretable modeling of the covariance structure. The within outcomes covariance matrix is specified for each marginal outcome using a linear combination of known matrices, while the joint covariance matrix is specified using the generalized Kronecker product \citep{martinez13,Bonat16}.

@@

When working with regression models, a common interest is to verify if the absence of a certain explanatory variable of the model would generate a loss in the adjustment. Thus, a conjecture of interest is to assess whether there is sufficient evidence in the data to state that a given explanatory variable has no effect on the response. This is done through hypothesis testing. In this context, three hypothesis tests are common: the likelihood ratio test, the Wald test, and the Lagrange multiplier test, also known as the score test. \cite{engle} describes the general formulation of the three tests. All of them are based on the likelihood function of the models.

@@

In the case of standard linear models, there are techniques such as analysis of variance (ANOVA), initially proposed by \cite{anova_fisher}. According to \cite{anova1}, ANOVA is one of the most widely used statistical methods to test hypotheses and it is present in virtually all introductory statistics materials. The objective of the technique is to evaluate the effect of each explanatory variables on the response. This is done by comparing models with and without each of the explanatory variables. Therefore, this procedure makes it possible to assess whether the removal of each of the variables generates a significantly worse model when compared to the model with the variable. For the multivariate case, the technique of analysis of variance (ANOVA) is extended to the multivariate analysis of variance \citep{manova}, MANOVA. Among the multivariate hypothesis tests already discussed in the literature for MANOVAs, we have the Wilks' $\lambda$ \citep{wilks}, Hotelling-Lawley trace \citep{lawley,hotelling}, Pillai trace \citep{pillai} and largest root of Roy \citep{roy}.

@@

Complementary to ANOVAs and MANOVAs are multiple comparison tests. Such procedures are used when the analysis of variance points to the existence of a significant effect of the parameters associated with a categorical variable, that is, there is at least one significant difference between the levels of a factor. Thus, the multiple comparison test is another procedure based on hypothesis testing, used to determine where these differences appear. For example, suppose there is a three-level categorical variable $X$ in the model that assumes the levels: A, B, or C. The analysis of variance will show if there is an effect of the variable $X$ in the model, that is, if the response values are associated with the levels of $X$, however this result will not tell us whether the response values differ from A to B, or from A to C, or whether B differs from C. To detect such differences, multiple comparison tests are used, some of the tests discussed in the literature are: the Dunnett, Tukey, student's t test (LSD), Scott-Knott test, among others. \cite{hsu1996multiple} discusses various procedures for the purpose of multiple comparisons. \cite{bretz2008multiple} presents procedures for multiple comparisons in linear models.

@@

It is important to note that the asymptotic normality of estimators of McGLMs was established by \cite{Bonat16}, and Wald testing for estimators which are asymptotically normal is a standard approach in asymptotic statistics due to the central limit theorem. So, in this work, our objective is to show how to proceed to perform Wald tests for McGLMs and verify if the expected properties are met.

@@

We present the dataset in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model and present the main results of the analysis of the data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations. The data sets and R codes are available online and can be assessed in \href{https://github.com/lineu96/paper-htmcglm}{https://github.com/lineu96/paper-htmcglm}.

@@

The evaluation of binge eating was based on the binge eating scale (BES), one of the most used tools to measure binge eating in research, with numerous studies proving its effectiveness, translated into Portuguese and validated for individuals with obesity and submitted to bariatric surgery. This is a 16-item Likert scale questionnaire, prepared according to the 3rd edition of the Diagnostic and Statistical Manual of Mental Disorders \citep{spitzer1980diagnostic} by \cite{gormally1982assessment}. The individuals were instructed to select the option that best represented their answer and the final score was the sum of the points for each item, this score ranging from 0 to 46.

To assess food addiction, the Food Addiction Scale (YFAS) was used, a questionnaire that seeks to detect symptoms of addictive eating behaviors. YFAS was based on substance dependence criteria from Diagnostic and Statistical Manual IV – Proofreading (DSM-IV-TR) \citep{segal2010diagnostic} and endorsed for highly processed foods. This questionnaire was developed by \cite{gearhardt2009preliminary} and is a combination of 25 Likert scale options and the evaluation option used was the number of addiction symptoms.

@@

\noindent where the $g_r()$ are standard link functions that relates the response to the predictor; $\boldsymbol{X}_r$ denotes a $N \times k_r$ design matrix; $\boldsymbol{\beta}_r$ denotes a  $k_r \times 1$ vector of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ denotes the generalized Kronecker product \citep{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

The power variance function characterizes the Tweedie family of distributions, it is given by $\vartheta\left(\cdot; p_r\right) = \mu^{p_r}_r$, in which some distributions stand out: Normal ($p $ = 0), Poisson ($p$ = 1), gamma ($p$ = 2) and Inverse Gaussian ($p$ = 3) \citep{Jorgensen87, Jorgensen97}.

The Poisson–Tweedie dispersion function \citep{Jorgensen15} is an option for responses that characterize counts. The dispersion function given by $\vartheta\left(\cdot; p\right) = \mu + \tau\mu^p$ where $\tau$ is the dispersion parameter. Thus, we have a rich class of models to deal with responses that characterize counts, since many important distributions appear as special cases, such as: Hermite ($p$ = 0), Neyman type A ($p$ = 1), negative binomial ($p$ = 2) and Poisson–inverse Gaussian (p = $3$).

@@

The dispersion matrix $\boldsymbol{\Omega({\tau})}$ describes the part of the covariance within each response variable that does not depend on the mean structure, ie, the correlation structure between the sample observations. Based on the ideas of \cite{Anderson73} and \cite{Pourahmadi00}, \cite{Bonat16} proposed to model the dispersion matrix through a matrix linear predictor combined with a covariance link function given by:

@@

Some possible covariance link functions are identity, inverse and exponential-matrix. The specification of the covariance link function is discussed by \cite{Pinheiro96} and it is possible to select combinations of matrices to obtain the most well-known models in the literature for longitudinal data, time series, spatial and spatiotemporal data. Further details are discussed by \cite{Demidenko13}.

@@

McGLMs are fitted based on the estimating function approach described in detail by \cite{Bonat16} and \cite{jorg04}. This subsection presents an overview of the algorithm and the asymptotic distribution of estimators based on estimating functions.

@@

To estimate the regression parameters, the following quasi-score function \citep{Liang86} is used,

$$
\begin{aligned}
  \psi_{\boldsymbol{\beta}}(\boldsymbol{\beta},
  \boldsymbol{\lambda}) = \boldsymbol{D}^\top
  \boldsymbol{C}^{-1}(\mathcal{Y} - \mathcal{M}),
\end{aligned}
$$

@@

To take into account the covariance between the vectors $\boldsymbol{\beta}$ and $\boldsymbol{\lambda}$, \cite{Bonat16} obtained the sensitivity and cross-variability matrices, denoted by $S_{\boldsymbol{ \lambda \beta}}$, $S_{\boldsymbol{\beta \lambda}}$ and $V_{\boldsymbol{\lambda \beta}}$, more details in \cite{Bonat16}. The joint sensitivity and variability matrices of $\psi_{\boldsymbol{\beta}}$ and $\psi_{\boldsymbol{\lambda}}$ are denoted by

@@

To solve the system of equations $\psi_{\boldsymbol{\beta}} = 0$ and $\psi_{\boldsymbol{\lambda}} = 0$, \cite{jorg04} proposed the modified Chaser algorithm, which is defined as

@@

Let’s consider now we are interested in to assess whether there is sufficient evidence to state that there is an effect of the explanatory variable $X_1$ on both responses simultaneously. 
In this case, we have to test 2 parameters: $\beta_{11}$, which associates $X_1$ with the first response and $\beta_{21}$, which associates $X_1$ with the second response. 
We can write the hypothesis as follows:

@@

However, in cases where the responses have identical linear predictors and the assumptions about the parameters do not change from response to response, an alternative specification of the procedure is to use the Kronecker product as used in \cite{plastica}.

@@

Based on the proposal to use the Wald test for McGLMs, we aim to propose in this article three different procedures for generating ANOVA and MANOVA tables for regression parameters, following the nomenclature types I, II and III. In addition, we also aim to propose a procedure similar to ANOVA and MANOVA to evaluate the dispersion parameters of a given model. In the case of ANOVAs, a table is generated for each response variable. For MANOVAs only one table is generated, therefore, in order to be able to perform MANOVAs, the model responses must be subject to the same linear predictor.

@@

When ANOVA shows a significant effect of a categorical variable, it is usually of interest to assess which of the levels differ from each other. For this, multiple comparison tests are used. In the literature there are several procedures to perform such tests, many of them described in \cite{hsu1996multiple}.

@@

To carry out a test of multiple comparisons, it is enough to select the desired contrasts in the lines of the matrix $\boldsymbol{K_1}$ and use these lines as a matrix for specifying the hypotheses of the Wald test. Finally, due to the amount of hypothesis testing in a multiple comparison procedure, correction of p-values is recommended, we used Bonferroni correction.

@@

We can opt for a type II multivariate analysis of variance to assess the importance of the variables in the problem. This result, presented in the \autoref{tab4}, points to the significant existence of the moment effect and the absence of the group effect, indicating that for both responses the metrics change over time but without change between groups.

@@

%%% Supplementary materials (if any)
%%% ------------------------------------------
\section*{Supplementary materials}
R code and data set are available at https://github.com/lineu96/paper-htmcglm
